%
% chapter.tex -- 
%
% (c) 2008 Prof Dr Andreas Mueller
%
\chapter{Hyperbolic partial differential equations
\label{chapter-hyperbolisch}}
\lhead{Hyperbolic equations}
\rhead{}
In this chapter we discuss the wave equation as a very prominent
example of a hyperbolic equation.
Just like in the parabolic case, the time coordinate has special
significance.
The wave equation also is a kind of equation of motion, but of second
order, and thus more similar to the equations we are used to from
mechanics.

While a change in initial conditions or boundary conditions 
influences the solution everywhere, the wave equation behaves
very differently in that such changes propagate with finite speed.
For each point in the domain there are points that can be affected
by a change of the value in the point and others that don't notice
the change.
The interfaces between these domains are an important aspect that
helps understand the solutions.

To get closer to a solution, we start with the one dimensional case
and we will find a solution as a superposition of waves that travel in
both directions.
We then embark on a study how initial conditions influence the solutions.
Asking which initial conditions lead to a well posed problem then
leads us to the concept of characteristics.

\input{9-hyperbolic/separation.tex}
\input{9-hyperbolic/waves.tex}
\input{9-hyperbolic/cauchy.tex}
\input{9-hyperbolic/characteristics.tex}
\input{9-hyperbolic/discontinuities.tex}
\input{9-hyperbolic/higherdim.tex}
\input{9-hyperbolic/boundary.tex}

\section{Summary}
\begin{enumerate}
\item
The solution surface of a hyperbolic partial differential equation can be
covered by characteristic strips.
\item
Discontinuities or other singularities propagate along characteristics.
\item
As for quasilinear partial differential equations of first order,
characteristics can be used to determine which boundary values or
normal derivatives need to be specified for a hyperbolic problem to
be well posed.
\end{enumerate}

