%
% pdgl2ord.tex -- Klassifikation linearer PDGL zweiter Ordnung
%
% (c) 2016 Prof Dr Andreas Mueller, Hochschule Rapperswil
%
\chapter{Partial differential equations of second order\label{chapter-2ordnung}}
\lhead{Second Order partial differential equations}
Linear partial differential equations of second order have the form
\begin{equation}
\sum_{i,j=1}^na_{ij}\frac{\partial^2}{\partial x_i\,\partial x_j} u
+
\sum_{i=1}^nb_i\frac{\partial}{\partial x_i} u+cu=f.
\label{operator2ordnung}
\end{equation}
All the examples in chapter
\ref{chapter:examples}
match this template.
In spite of the superficial similarities, the solutions to these problems
exhibited completely different behaviour.
\begin{itemize}
\item The wave equation has solutions that propagate with finite 
velocity.
A change in initial conditions is noticed in more distant points of
the domain only after some delay.
\item
Changes in the boundary conditions of the heat equation affect the solution
for later times instantaneously, they propagate with infinite speed through
the domain, but they don't affect the past.
\item
Changing the initial conditions of the poisson problem
$\Delta \varphi=f$,
changes the solution immediately everyhwere in the domain.
\end{itemize}
In this chapter we will classify partial differential equations of
second order.
The goal is to show that any linear partial differential differential
equation of second order falls into one of those three categories.

\input{6-2ord/standard.tex}
\input{6-2ord/types.tex}
\input{6-2ord/canonical.tex}
\input{6-2ord/splitting.tex}

