%
% elliptic.tex -- 
%
% (c) 2019 Prof Dr Andreas Mueller
%
\chapter{Elliptic partial differential equations
\label{chapter:elliptic}}
In this chapter we study the solutions of elliptic partial differential
equations and discover the maximum-minimum principle as well as the
the mean value property of harmonic functions.
At the end of the chapter we describe Green's function, which allows
to give a formula for the solution of a boundary value problem.

\input{7-elliptic/intro.tex}
\input{7-elliptic/unique.tex}
\input{7-elliptic/inverse.tex}
\input{7-elliptic/1dim.tex}
\input{7-elliptic/general.tex}
\input{7-elliptic/mean.tex}
\input{7-elliptic/generalizations.tex}

\section{Summary}
\rhead{Summary}
\begin{enumerate}
\item
The differential equation $\Delta u=f$ is the prototypical elliptic
partial differential equation.
\item
Solutions of a homogeneous elliptic partial differential equation,
in particular the equation $\Delta u=0$, satisfy the maximum principle:
minimum and maximum of $u$ lie on the boundary of the domain.
\item
The solutions of elliptic partial differential equation with boundary
values along the boundary of bounded domain are unique.
\item
A well posed elliptic differential equation has a Green's function,
i.~e.~a solution of the differential equation
$\Delta G(x,\xi)=\delta_\xi$ for every point in $\xi\in\Omega$
with homogeneous boundary conditions.
\item
The Green's function can be used to solve the inhomogeneous equation
by means of an integral over the domain.
\item
Harmonic functions are solutions of the equation $\Delta u=0$.
\item
Harmonic functions have the mean value property: the value of a harmonic
function in a point is the mean of the function values on a sphere around
the point.
\end{enumerate}


