Solve the partial differential equation
\[
\frac{\partial^2}{\partial x^2}u+\frac{\partial^2}{\partial y^2}u
+2\frac{\partial }{\partial y}u+u=0
\]
on the domain $0 < x < \pi$, $y > 0$ with boundary conditions
\[
u(0,y)=u(\pi,y)=0
\quad
\quad
u(x,0)=\sin x,\quad \frac{\partial }{\partial n}u(x,0)=\sin 3x
\]

\begin{loesung}
The separation ansatz $u(x,y)=X(x)Y(y)$ gives
\begin{align*}
X''(x)Y(y)+X(x)Y''(y)+2X(x)Y'(y)+X(x)Y(y)&=0
\\
\frac{Y''(y)+2Y'(y)+Y(y)}{Y(y)}
&=
-\frac{X''(x)}{X(x)}.
\end{align*}
We derive the ordinary differential equations
\begin{align*}
X''(x)&=-\lambda X(x)\\
Y''(y)+2Y'(y)+(1-\lambda)Y(y)&=0.
\end{align*}
For positive $\lambda$, the first equation has the solution
$\sin\sqrt{\lambda}x$, which satisfies the boundary condition for $x=0$,
but for $x=\pi$ it does so only if $\sqrt{\lambda}$ is an integer, so
$\lambda=n^2$.
For negative $\lambda$, there is no solution that respects the bounary
condition.
Therefore the second equation reduces to
\[
Y''(y)+2Y'(y)+(1-n^2)Y(y)=0.
\]
This is a linear differential equation with constant coefficients,
which can be solved using one of the well known methods.
The characteristic polynomial is
\[
\omega^2+2\omega+(1-n^2)=0,
\]
it has the zeros
\[
\omega_{\pm}=-1\pm{\sqrt{1-(1-n^2)}}=-1\pm n.
\]
The general solution is
\[
u_n(x,y)=(a_{n,+}e^{(-1+n)y}+a_{n,-}e^{(-1-n)y})\sin nx.
\]
The exponential factors can be more elegantly expressed by hyperbolic
functions:
\[
u_n(x,y)=(a_n\cosh ny+b_n\sinh ny)e^{-y}\sin nx.
\]
Indeed, we can verify that these functions are in fact solutions.
Its derivatives are
\begin{align*}
\frac{\partial}{\partial x}u_n(x,y)
&=
n (a_n\cosh ny+b_n\sinh ny)e^{-y}\cos nx
\\
\frac{\partial^2}{\partial x^2}u_n(x,y)
&=
-n^2 (a_n\cosh ny+b_n\sinh ny)e^{-y}\sin nx
\\
\frac{\partial}{\partial y}u_n(x,y)
&=
-(a_n\cosh ny+b_n\sinh ny)e^{-y}\sin nx
+
n(a_n\sinh ny+b_n\cosh ny)e^{-y}\sin nx
\\
\frac{\partial^2}{\partial y^2}u_n(x,y)
&=
(a_n\cosh ny+b_n\sinh ny)e^{-y}\sin nx
-2n(a_n\sinh ny+b_n\cosh ny)e^{-y}\sin nx
\\
&\qquad
+n^2(a_n\cosh ny+b_n\sinh ny)e^{-y}\sin nx.
\end{align*}
By substituting this in the differential equation, we get
\begin{align*}
-n^2 (a_n\cosh ny+b_n\sinh ny)e^{-y}\sin nx&\\
+(a_n\cosh ny+b_n\sinh ny)e^{-y}\sin nx&\\
-2n(a_n\sinh ny+b_n\cosh ny)e^{-y}\sin nx&\\
+n^2(a_n\cosh ny+b_n\sinh ny)e^{-y}\sin nx&\\
+2(
-(a_n\cosh ny+b_n\sinh ny)e^{-y}\sin nx&\\
+
n(a_n\sinh ny+b_n\cosh ny)e^{-y}\sin nx
)&\\
+
(a_n\cosh ny+b_n\sinh ny)e^{-y}\sin nx
&=0.
\end{align*}
In this form, the initial conditions are a bit more transparent to
satifisfy, because $\sinh n0=0$ and $\cosh n0=1$.
The normal derivative on the boundary $y=0$ is the derivative in the
direction $y$, so
\begin{align*}
u_n(x,0)&=a_n\sin nx
\\
\frac{\partial}{\partial n}u_n(x,0)&=(-a_n+nb_n)\sin nx.
\end{align*}
So from the boundary values for $y=0$ we can determine $a_n$ and from
the values of the normal derivative for $y=0$ we get $b_n$.

In the present case, $a_1=1$ is the only $a$-coefficient that does
not vanish.
To satisfy the boundary conditions for the normal derivative, we need $n=$.
It follows that
\begin{align*}
a_1&=1&&\\
-a_1+1\cdot b_1&=0&\Rightarrow\quad b_1&=1\\
a_3&=0&&\\
-a_3+3b_3&=1&\Rightarrow\quad b_3&=\frac13.
\end{align*}
So the solution we are looking for is
\[
u(x,y)=e^{-y}(\sin x \cosh y +
\sin x\sinh y
+
\frac13\sin 3x \sinh 3y),
\]
which we can also verify e.~g.~using a computer algebra system.
\end{loesung}

