%
% classification.tex
% 
% (c) 2019 Prof Dr Andreas Mueller
%
\chapter{Terminology and Notation
\label{chapter:terminology-and-notation}}
\lhead{Terminology and notation}
Partial differential equations are equations for an unknown function of
several variables, linking values of the function to its partial
derivatives.
In this chapter, we write the unknown function always as $u(x_1,\dots,x_n)$
and the independent variables as $x_1,\dots,x_n$.
\[
u\colon \mathbb R^n\to\mathbb R:(x_1,\dots,x_n)\mapsto u(x_1,\dots,x_n).
\]
We will abbreviate points of the domain of definition of such a function
as $x=(x_1,\dots,x_n)$.
We have to clarify the following aspects:
\begin{compactenum}
\item What is a partial differential equation exactly?
\item Where is it defined?
\item How do we formulate boundary conditions and initial conditions?
\item What types of partial equations are there?
\end{compactenum}

\input{2-classification/equations.tex}
\input{2-classification/domains.tex}
\input{2-classification/conditions.tex}
\input{2-classification/solutions.tex}
\input{2-classification/types.tex}
\input{2-classification/notation.tex}

\section{Summary}
\begin{enumerate}
\item
A partial differential equation is specified by a function $F$
of the independent variables, of the function $u$ and of placeholders
for the partial derivatives of $u$.
\item
The order of a partial differential equation is the order of the
highest order partial derivative.
\item
In two dimensions, a partial differential equation of first order can
always be described by a function $F(x,y,u,p,q)$.
The variables $p$ and $q$ are place holders for the first partial
derivatives of $u$:
\[
F\biggl(
x,y, u,
\frac{\partial u}{\partial x},
\frac{\partial u}{\partial y}
\biggr)=0.
\]
\item
A partial differential equation can always be reduced to a system
of partial differential equations of first order.
(\ref{klassifikation:umwandlung}).
\item
\index{linear}
\index{partial differential equation!linear}
Linear partial differential equations are have $F$ depend on the function 
$u$ and the derivatives only linearly.
In two dimension, this means that $F(x,y,u,p,q)$ is linear in
$u$, $p$ and $q$.
\item
A {\em quasilinear} partial differential equation is linear in all
\index{quasilinear}
\index{partial differential equation!quasilinear}
partial derivatives, but not necessarily in the function $u$ itself.
(\ref{klassifikation:quasilinear}).
\item
The domain $\Omega$ where a differential equation is defined
has a significant influence on the solution
(\ref{klassifikation:gebiete}).
\item
The solution becomes well defined only after suitable boundary conditions
have been specified
(\ref{klassifikation:randbedingungen}). 
\item
For partial differential equations of first order we expect to specify
values of the function on part of the boundary (Cauchy problem).
\item
For partial differential equations of second order we expect to
specify values (Dirichlet boundary conditions)
and or normal derivatives (Neumann boundary conditions).
\item
A solution of a partial differential equation is a function defined
on the closure of the domain, that satisfies the differential equation
in the domain and boundary conditions on the boundary.
(\ref{klassifikation:loesung}).
\end{enumerate}
